---
title: "Telomere length assay report"
author: "Selene Banuelos"
format:
  html:
    code-fold: true
editor: visual
---

This is a report on telomere length-related data that is currently available for 
PEARLS participants. For Run 1 and Run 2 of the telomere assay, this report
contains information about any missing telomere or 36B4 data, DNA quality check,
and plate contamination for all T5 buccal samples. Recommendations for samples
to assay next are included at bottom of report. 

## Setup

```{r message = FALSE}
# load libraries
library(dplyr)
library(tidyr)

# import telomere/36B4/buccal DNA QC data
tel_data <- read.csv('data-processed/merged-runs-tel-36b4.csv')

# telomere data availability
avail <- read.csv('data-processed/tel-avail-reruns-T5.csv')
```

## Run 1

### Missing data
Check if any samples are missing telomere-specific or 364b-specific probe data

```{r}
avail %>%
  # check run 1 data
  filter(run == 1) %>%
  # look at samples with data available for either telomere or 36B4
  filter(data_avail == 1) %>%
  # keep only relevant variables
  select(specimenid, Target, run, data_avail) %>%
  # remove triplicatesS
  distinct(.) %>%
  # make data wider to check data availability across both probes
  pivot_wider(id_cols = c('specimenid', 'run'),
              names_from = Target,
              names_glue = 'avail_{Target}',
              values_from = data_avail) %>%
  # make inidcator of data missing eitehr for telomere or 36B4-specific probes
  mutate(missing_either = case_when(is.na(avail_TEL) | is.na(avail_36B4) ~ 1,
                                    .default = 0)
         ) %>%
  # check any samples that are missing data in either
  filter(missing_either == 1) %>%
  nrow(.)
```

All samples have data for both probes.

### DNA QC

Generate list of samples that were included in Run 1 but did not pass DNA QC metrics. To pass DNA QC, DNA concentration must be greater than or equal to 5ng/ul and 260/280 ratio between 1.7-2.0.

```{r}
not_pure1 <- tel_data %>%
  # select rows from run 1
  filter(run == 1) %>%
  # remove standards, positive controls, no template controls
  filter(!grepl('Std|Pos Ctrl|NTC', Content)) %>%
  # keep samples that did not pass QC
  filter(dna_qc_passed == 0) %>%
  # get vector of sample IDs
  pull(Sample) %>%
  # remove triplicates
  unique(.)

print(not_pure1)
```

It may be worth checking these samples again on Nanodrop. It appears that these had sufficient DNA concentration but 260/280 ratio does not fall in an acceptable range for DNA or 260/280 ratio is missing. See below:

```{r}
tel_data %>%
  # only show samples that did not pass DNA QC
  filter(Sample %in% not_pure1) %>%
  # keep only relevant samples
  select(Sample, os_dna_conc, purity, dna_qc_passed) %>%
  # remove triplicates
  distinct(.)
```

### Contamination

Check if any no template control (NTC) wells in plates included in Run 1 have Cq/Ct value\<35

```{r}
tel_data %>%
  # select rows from run 1
  filter(run == 1) %>%
  # indicator variable of NTC Cq/Ct value<35
  filter(ntc_cont == 1) %>%
  nrow(.)
```

Looks good, no signs of contamination in Run 1 in either telomere or 36B4 plates.

## Run 2

### Missing data
Check if any samples are missing telomere-specific or 364b-specific probe data

```{r}
avail %>%
  # check run 1 data
  filter(run == 2) %>%
  # look at samples with data available for either telomere or 36B4
  filter(data_avail == 1) %>%
  # keep only relevant variables
  select(specimenid, Target, run, data_avail) %>%
  # remove triplicatesS
  distinct(.) %>%
  # make data wider to check data availability across both probes
  pivot_wider(id_cols = c('specimenid', 'run'),
              names_from = Target,
              names_glue = 'avail_{Target}',
              values_from = data_avail) %>%
  # make inidcator of data missing eitehr for telomere or 36B4-specific probes
  mutate(missing_either = case_when(is.na(avail_TEL) | is.na(avail_36B4) ~ 1,
                                    .default = 0)
         ) %>%
  # check any samples that are missing data in either
  filter(missing_either == 1) %>%
  nrow(.)
```

All samples from Run 2 have telomere and 36B4-specific probe data.

### DNA QC

Generate list of samples that were included in Run 2 but did not pass DNA QC metrics. To pass DNA QC, DNA concentration must be greater than or equal to 5ng/ul and 260/280 ratio between 1.7-2.0.

```{r}
not_pure2 <- tel_data %>%
  # select rows from run 2 that were not reruns from run 1
  filter(run == 2 & rerun == 0) %>%
  # remove standards, positive controls, no template controls
  filter(!grepl('Std|Pos Ctrl|NTC', Content)) %>%
  # keep samples that did not pass QC
  filter(dna_qc_passed == 0) %>%
  # get vector of sample IDs
  pull(Sample) %>%
  # remove triplicates
  unique(.)

print(not_pure2)
```

It may be worth checking these samples again on Nanodrop. It appears that these had sufficient DNA concentration but 260/280 ratio does not fall in an acceptable range for DNA or 260/280 ratio is missing. See below:

```{r}
tel_data %>%
  # only show samples that did not pass DNA QC
  filter(Sample %in% not_pure2) %>%
  # keep only relevant samples
  select(Sample, os_dna_conc, purity, dna_qc_passed) %>%
  # remove triplicates
  distinct(.)
```

### Contamination

Check if any telomere or 36B4 plates from Run 2 have signs of contamination (Cq/Ct value\<35)

```{r}
cont_plates <- tel_data %>%
  # select rows from run 1
  filter(run == 2) %>%
  # indicator variable of NTC Cq/Ct value<35
  filter(ntc_cont == 1) %>%
  # get vector of file names corresponding to contaminated plates
  pull(file_name) %>%
  # remove triplicates
  unique(.)

print(cont_plates)
```

`{r} length(cont_plates)` plates have signs of contamination based on the NTC wells.

Get list of samples that were included in these `{r} length(cont_plates)` contaminated plates and check to see if any were also included in Run 1

```{r}
# get vector of specimen IDs of samples on contaminated plates that were only assayed once
once <- tel_data %>%
  # remove standards, positive controls, no template controls
  filter(!grepl('Std|Pos Ctrl|NTC', Content)) %>%
  # filter for samples from contaminated plates
  filter(ntc_cont == 1) %>%
  # filter for samples run only once
  filter(rerun == 0) %>%
  select(Target, Sample, ntc_cont, dna_qc_passed) %>%
  # remove triplicates
  distinct(.) %>%
  # make data wider to check for contamination across telomere and 36B4
  pivot_wider(id_cols = c('Sample', 'dna_qc_passed'),
               names_from = Target,
               names_prefix = 'contaminated_',
               values_from = ntc_cont)

# get vector of specimen IDs of samples on contaminated plates that were assayed more than once
rerun <- tel_data %>%
    # remove standards, positive controls, no template controls
  filter(!grepl('Std|Pos Ctrl|NTC', Content)) %>%
  # filter for samples from contaminated plates
  filter(ntc_cont == 1) %>%
  # filter for samples that were rerun
  filter(rerun == 1) %>%
  select(Target, Sample, ntc_cont, dna_qc_passed) %>%
  # remove triplicates
  distinct(.) %>%
  # make data wider to check for contamination across telomere and 36B4
  pivot_wider(id_cols = c('Sample', 'dna_qc_passed'),
               names_from = Target,
               names_prefix = 'contaminated_',
               values_from = ntc_cont)
```

Across all contaminated plates in Run 2, there were `{r} nrow(rerun)` samples that were also included in Run 1. Since Run 1 did not show signs of contamination in any plate, data from that run can be used for these samples.

Across all contaminated plates in Run 2, there were `{r} nrow(once)` samples that were only included in Run 2. From these `{r} nrow(once)` samples:

-    `{r} filter(once, contaminated_TEL == 1 & contaminated_TEL == 1) %>% nrow(.)` had telomere and 36B4 data from contaminated plates

-   `{r} filter(once, is.na(contaminated_TEL) | is.na(contaminated_36B4)) %>% nrow(.)` had only 36B4 data from contaminated plates (telomere data for these samples came from plates with NTC Cq\>35)

## Samples to run next

```{r}
need_data <- avail %>%
  # keep samples that weren't included in Runs 1 or 2
  filter(data_avail == 0) %>%
  # save vector of specimen ID
  select(specimenid) %>%
  distinct(.) %>%
  pull(specimenid)
```

### Sample set A 
The following `{r} length(need_data)` samples were originally selected for Hector to run since they had not been included in either Run 1 or Run 2:

`{r} need_data`

### Sample set B 
The following `{r} nrow(once)` samples were only included in Run 2 and have telomere and/or 36B4 data from contaminated plates:

`{r} once$Sample`

### Sample set C
In addition to those samples, you may want to consider re-measuring 260/280 ratios to re-assess purity of the samples below that were already run in either Run 1 or Run 2:

-   `{r} length(not_pure1)` samples from Run 1: `{r} not_pure1`

-   `{r} length(not_pure2)` samples from Run 2: `{r} not_pure2`
